---
title: "DC-TAP-seq indirect effect sizes"
author: "Andreas R. Gschwind"
date: "July 23, 2026"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
---

```{r setupDocument, include=FALSE}
# set output html chunk options
knitr::opts_chunk$set(warning = FALSE, message = FALSE)

# suppress dplyrs summarize info
options(dplyr.summarize.inform = FALSE)
```

```{r requiredPackages, warning=FALSE, message=FALSE}
library(tidyverse)
library(cowplot)
```

## Goal
Plot effect sizes of indirect effects in K562 and WTC11 DC-TAP-seq datasets.
```{r}
# path to indirect effects analysis output (adjust to reproduce)
indirect_effects_dir <- "/oak/stanford/groups/engreitz/Users/agschwin/distal_regulation_paper/CRISPR_indirect_effects_FDR_check/results"

# path to Gasperini et al indirect effects analysis output (adjust to reproduce)
gasp_dir <- "/oak/stanford/groups/engreitz/Users/agschwin/distal_regulation_paper/ENCODE_CRISPR_data/results/Gasperini2019"
```

***

## Used data
Sceptre and MAST output used to calculated indirect effect rates across all datasets is used to
analyze effect sizes of detected significant indirect effects.
```{r}
# all output files for sceptre-based trans-acting effect tests
scpt_datasets <- c("K562_DC_TAPseq", "WTC11_DC_TAPseq", "Klann2021", "Xie2019", "Morris2023_v1", "Morris2023_v2")
scpt_files <- file.path(indirect_effects_dir, scpt_datasets, "results_run_discovery_analysis.rds")
names(scpt_files) <- scpt_datasets

# load all sceptre result files
sceptre_trans_results <- lapply(scpt_files, FUN = readRDS)

# all output files for sceptre-based cis analysis
cis_files <- file.path(indirect_effects_dir, scpt_datasets, "annotated_cis_results.tsv.gz")
names(cis_files) <- scpt_datasets

# load all cis analysis results
sceptre_cis_results <- lapply(cis_files, FUN = read_csv, show_col_types = FALSE)

# gasperini et al trans and cis result files
gasp_trans_file <- file.path(gasp_dir, "trans_effects/output_trans_effects_MAST_perCRE.tsv.gz")
#gasp_cis_file <- file.path(gasp_dir, "diff_expr/output_MAST_perCRE.tsv.gz")
gasp_encode_file <- file.path(gasp_dir, "../ENCODE/ENCODE_Gasperini2019_0.13gStd_MAST_perCRE_GRCh38.tsv.gz")

# load gasperini files
gasp_trans_results <- read_tsv(gasp_trans_file, show_col_types = FALSE)
#gasp_cis_results <- read_tsv(gasp_cis_file, show_col_types = FALSE)
gasp_encode <- read_tsv(gasp_encode_file, show_col_types = FALSE)

# load gene universe file
gene_univ <- read_tsv("/oak/stanford/groups/engreitz/Users/agschwin/distal_regulation_paper/CRISPR_benchmarks/resources/genome_annotations/CollapsedGeneBounds.hg38.TSS500bp.bed",
                      show_col_types = FALSE)
```

***

## Extract effect sizes for sigificant indirect hits
Filter trans differential expression results using the same filters applied ot calculate indirect
effect rates and extract indirect effect sizes.

### Sceptre based results
```{r}
# helper function to extract significant indirect effects for one dataset
get_indirect_effects <- function(trans_results, cis_results, gene_univ) {
  
  # get significance threshold for cis effects
  pval_threshold_cis <- cis_results %>% 
    filter(significant == TRUE) %>% 
    pull(p_value) %>% 
    max()
  
  # add significance and regulated column to trans-effect results based on cis significance threshold
  trans_results <- trans_results %>% 
    mutate(significant = p_value <= pval_threshold_cis,
           regulated_negative = significant & log_2_fold_change < 0,
           regulated_positive = significant & log_2_fold_change >= 0)
  
  # get all valid elements from cis results
  valid_elements <- cis_results %>% 
    filter(valid_element == TRUE) %>% 
    pull(grna_target) %>% 
    unique()
  
  # add a valid connection column to trans-results based on whether perturbation is a valid element
  trans_results <- mutate(trans_results, valid_connection = grna_target %in% valid_elements)
  
  # filter trans results for valid connections and only retain data on gene universe
  trans_results_filt <- trans_results %>% 
    filter(valid_connection == TRUE, pass_qc == TRUE) %>% 
    filter(response_id %in% gene_univ$Ensembl_ID)
  
  # extract significantly regulated indirect effects
  indirect_effects <- trans_results_filt %>% 
    filter(regulated_negative == TRUE | regulated_positive == TRUE) %>% 
    mutate(effect_direction = case_when(
      regulated_negative & regulated_positive ~ "both (ERROR)",
      regulated_negative ~ "negative",
      regulated_positive ~ "positive",
      TRUE ~ NA_character_
    ))
  
  return(indirect_effects)
}

# helper function to convert log-fold changes to percentage changes
logFC_to_pctChange <- function(logFC, base = exp(1)) {
  pct_change <- base ^ logFC - 1
  return(pct_change)
}
```

```{r}
# extract significant indirect effects for both dataset
indirect_effects <- mapply(FUN = get_indirect_effects, sceptre_trans_results, sceptre_cis_results,
                           MoreArgs = list(gene_univ = gene_univ), SIMPLIFY = FALSE)

# combine into one table and convert effect sizes from log2FC to percent change in expression
indirect_effects <- indirect_effects %>% 
  bind_rows(.id = "dataset") %>% 
  mutate(pct_change = logFC_to_pctChange(log_2_fold_change, base = 2))
```

### MAST based results (Gasperini dataset)
```{r}
# get valid enhancers for gasperini dataset based on ENCODE output
enh_filter <- c("overlaps potential promoter", "TSS targeting guide(s)")
gasp_valid_enh <- gasp_encode %>% 
  filter(!ValidConnection %in% enh_filter) %>% 
  pull(PerturbationTargetID)

# add ValidConnection column to gasperini trans-results
gasp_trans_results <- gasp_trans_results %>% 
  mutate(pert_id = paste0(pert_chr, ":", pert_start, "-", pert_end, ":.")) %>% 
  mutate(ValidConnection = if_else(pert_id %in% gasp_valid_enh, true = "TRUE",
                                   false = "Invalid enhancer"))

# filter trans results based on ValidConnection
gasp_trans_results <- filter(gasp_trans_results, ValidConnection == "TRUE")

# extract significantly regulated indirect effects
gasp_trans_results <- gasp_trans_results %>% 
  filter(regulated_negative == TRUE | regulated_positive == TRUE) %>% 
  mutate(effect_direction = case_when(
    regulated_negative & regulated_positive ~ "both (ERROR)",
    regulated_negative ~ "negative",
    regulated_positive ~ "positive",
    TRUE ~ NA_character_
  ))

# convert effect sizes from log2FC to percent change in expression
gasp_trans_results <- gasp_trans_results %>% 
  mutate(pct_change = logFC_to_pctChange(logFC, base = exp(1)))
```

### Combine into one table
```{r}
# select relevant columns and combine from sceptre based trans results
indirect_effects <- indirect_effects %>% 
  select(dataset, perturbation = grna_target, gene = response_id, regulated_negative,
         regulated_positive, effect_direction, pct_change)

# select relevant columns and combine from gasperini trans results
gasp_trans_results <- gasp_trans_results %>% 
  mutate(dataset = "Gasperini2019") %>% 
  select(dataset, perturbation, gene, regulated_negative, regulated_positive, effect_direction,
         pct_change)

# combine into one table
indirect_effects <- bind_rows(indirect_effects, gasp_trans_results)
```

***

## Plot indirect effect sizes

Group data for plots: 1 = K562 DC-TAP-seq, 2 = WTC11 DC-TAP-seq, 3 = Other datasets.
```{r}
# redefine datasets to group data for plots
indirect_effects <- indirect_effects %>% 
  mutate(dataset_plot = case_when(
    dataset == "K562_DC_TAPseq" ~ "K562",
    dataset == "WTC11_DC_TAPseq" ~ "WTC11",
    TRUE ~ "Other datasets"
  )) %>% 
  mutate(dataset_plot = factor(dataset_plot, levels = c("K562", "WTC11", "Other datasets")))
```

```{r}
# label outliers in each plotted distribution for filtering to prevent long tails on violin plots
indirect_effects <- indirect_effects %>%
  group_by(dataset_plot, effect_direction) %>% 
  mutate(outlier = !between(pct_change, 
         quantile(pct_change, 0.25) - 1.5 * IQR(pct_change),
         quantile(pct_change, 0.75) + 1.5 * IQR(pct_change)))

# calculate percentage of outlier points removed per distribution
indirect_effects <- indirect_effects %>% 
  mutate(total_pairs = n(),
         outlier_pairs = sum(outlier),
         pct_outliers = outlier_pairs / total_pairs)

# filter out outlier points
indirect_effects_filt <- filter(indirect_effects, outlier == FALSE)

# create labels with number of hits for plots
indirect_effects_filt <- indirect_effects_filt %>% 
  mutate(filt_pairs = n()) %>% 
  mutate(label = paste0(effect_direction, "\n(n = ", filt_pairs, " trans-pairs,\n", outlier_pairs,
                        " outliers removed)"))
```


Plot effect sizes for negative and positive indirect effects as violin plots.
```{r, fig.height=4, fig.width=11}
# plot effect sizes of indirect effects as violin plots
ggplot(indirect_effects_filt, aes(x = label, y = pct_change, fill = effect_direction)) +
  facet_wrap(~dataset_plot, drop = TRUE, scales = "free") +
  geom_hline(yintercept = 0, linetype = "dashed", linewidth = 0.5) +
  geom_violin() +
  geom_boxplot(color = "black", fill = NA, outlier.shape = NA, width = 0.2) +
  labs(y = "CRISPR effect size\n(% change in expression)", x  = "Indirect effect direction") +
  scale_fill_manual(values = c("firebrick3", "steelblue")) +
  scale_y_continuous(limits = range(indirect_effects_filt$pct_change), labels = scales::percent) +
  theme_classic() +
  theme(legend.position = "none", strip.background = element_blank())

# save plot to pdf
ggsave(file = "plots/indirect_effect_sizes_violin.pdf", height = 4, width = 11, create.dir = TRUE)
```

***

## Session information
This document was produced using following packages:
```{r sessionInfo}
sessionInfo()
```
