---
title: "DC-TAP-seq direct vs indirect effects supplementary analyses"
author: "Andreas R. Gschwind"
date: "July 23, 2026"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
---

```{r setupDocument, include=FALSE}
# set output html chunk options
knitr::opts_chunk$set(warning = FALSE, message = FALSE)

# suppress dplyrs summarize info
options(dplyr.summarize.inform = FALSE)
```

```{r requiredPackages, warning=FALSE, message=FALSE}
library(tidyverse)
library(cowplot)
```

## Goal
Analyze direct vs. indirect effects across datasets and plot modeled direct effect rates as function
of distance to TSS.
```{r}
# adjust path to indirect effects analysis output as needed to reproduce results
indirect_effects_dir <- "/oak/stanford/groups/engreitz/Users/agschwin/distal_regulation_paper/CRISPR_indirect_effects_FDR_check/results"
```

***

## Used data
Estimated cis and trans effect rates, as well as predicted direct rates and probability of direct vs
indirect effects are used to plot direct and indirect rates.
```{r}
# load cis and trans effect rates
cis_rates <- read_tsv(file.path(indirect_effects_dir, "cis_positive_hit_rates.tsv"),
                      show_col_types = FALSE)
trans_rates <- read_tsv(file.path(indirect_effects_dir, "trans_positive_hit_rates.tsv"),
                        show_col_types = FALSE)

# load direct effect rates per dataset and averaged across datasets
direct_rates_datasets <- read_tsv(file.path(indirect_effects_dir, "direct_effect_models/direct_rates_per_datasets.tsv"),
                                  show_col_types = FALSE)
direct_rates_average  <- read_tsv(file.path(indirect_effects_dir, "direct_effect_models/direct_rates_average_across_datasets.tsv"),
                                  show_col_types = FALSE)

# load direct effect models
models <- readRDS(file.path(indirect_effects_dir, "direct_effect_models/direct_effects_model.rds"))

# load dataset ids
dataset_ids <- read_tsv(file.path(indirect_effects_dir, "../config/crispr_dataset_ids.tsv"),
                        show_col_types = FALSE)

# add dataset ids to use in plots to cis, trans and direct rates
dataset_ids_rates <- select(dataset_ids, dataset, dataset_plots)
cis_rates <- left_join(cis_rates, dataset_ids_rates, by = "dataset")
trans_rates <- left_join(trans_rates, dataset_ids_rates, by = "dataset")
direct_rates_datasets <- left_join(direct_rates_datasets, dataset_ids_rates, by = "dataset")
```

***

## Model direct effect rate
The cis positive hit rate is a combination of the direct and indirect hit rate. The expected
indirect hit rate in these CRISPR experiments was estimated by the trans positive hit rate. We can
now calculate the estimated direct effect hit rates by subtracting the indirect effects rate
(trans hit rate) from the cis positive hit rate for each distance bin.

Some distance bins do have a lower (or 0) cis hit rate in some datasets due to the overall sparsity
of positive hits in these experiments. Thus the resulting negative direct hit rate will can be
considered an artifact.

To model the direct hit rate as a function of distance to TSS, we average the calculated direct
effects hit rate across all dataset per distance bin and fit a powerlaw model. The following plots
show the powerlaw fit visualized on the direct effect hit rates averaged across datasets.
```{r}
# plot direct effects model fit
plot_powerlaw_model_fit <- function(rate_type = c("all", "negative", "positive"),
                                    models, direct_rates_average, log = FALSE, ylim = NA) {
  
  rate_type <- match.arg(rate_type)
  
  # get direct rates for given rate type
  rate_type_data <- paste0("direct_rate_", rate_type)  # rate type in data and model list
  direct_rates_avg_type <- filter(direct_rates_average, type == rate_type_data)

  # get model coefficients for given rate type
  model_type <- models[[rate_type_data]]
  a <- exp(coef(model_type)[1])
  b <- coef(model_type)[2]
  
  # plot model fit vs average direct rate per dataset
  p <- ggplot(direct_rates_avg_type, aes(x = dist_to_tss, y = direct_rate)) +
    geom_point(color = "steelblue") +
    stat_function(fun = function(x) a * x^b, color = "black") +
    labs(title = str_to_sentence(paste("Direct hit rate", rate_type, "effects")),
         x = "Distance to TSS (bp)", y = "Direct hit rate") +
    scale_x_continuous(labels = scales::label_scientific(digits = 1)) +
    scale_y_continuous(limits = c(NA, ylim)) +
    theme_classic() +
    theme(text = element_text(size = 13))
  
  # transform plots to log space if specified
  if (log == TRUE) {
    p <- p +
      scale_x_log10() +
      scale_y_log10(limits = c(NA, ylim)) +
      annotation_logticks()
  }
  
  return(p)
  
}
```

Plot effect rates in log space. Note that this introduces infinite values in the direct hit rates
because some distance bins have no positives due to overall data sparsity.
```{r, fig.height=4, fig.width=8}
# plot all model fits
rate_types <- c(negative = "negative", positive = "positive")
model_fit_plots <- lapply(rate_types, FUN = plot_powerlaw_model_fit, models = models,
                          direct_rates_average = direct_rates_average,
                          ylim = 0.2, log = TRUE)

# arrange all plots into on figure
p1 <- plot_grid(plotlist = model_fit_plots, nrow = 1)
p1
```

***

## Predicted direct effects rate
Using the powerlaw model, we can predict the direct effects rate as function of distance to TSS. The
following plots show the predicted direct effect rates as function of distance to TSS for both
negative and positive effects on expression. Empirically calculated indirect effect rates are
overlayed as horizontal lines. 
```{r}
# function to predict the direct hit rate for a given distance
predict_direct_hit_rate <- function(models, new_data, type) {
  
  # get model for specified type
  type <- paste0("direct_rate_", type)
  fit <- models[[type]]
  
  # predict direct hit rate for provided distances and exponentiate to get back to original scale
  predicted_log_direct_rate <- predict(fit, newdata = new_data)
  predicted_direct_rate <- exp(predicted_log_direct_rate)
  
  # cap predicted values at 1, because the rate can't be higher than that and add to input data
  new_data$predicted_direct_rate <- pmin(predicted_direct_rate, 1)
  
  return(new_data)
  
}
```

```{r, fig.height=4, fig.width=10}
# convert trans rates to long format and calculate average indirect rate across datasets
trans_rates <- trans_rates %>% 
  select(dataset_plots, all = positive_rate_significant, negative = positive_rate_negative,
         positive = positive_rate_positive) %>% 
  pivot_longer(cols = -dataset_plots, names_to = "effect_direction", values_to = "indirect_rate") %>% 
  filter(effect_direction %in% rate_types)

# predict direct effects rate from 1bp to 1Mb distance to TSS and combine into one table
direct_effects_1mb <- lapply(rate_types, FUN = predict_direct_hit_rate, models = models, 
                             new_data = tibble(dist_to_tss = 1:1e6)) %>% 
  bind_rows(.id = "effect_direction")
```

```{r, fig.height=3, fig.width=8}
# prettier effect directions for plots
direct_effects_1mb <- direct_effects_1mb %>% 
  mutate(effect_direction = paste(str_to_title(effect_direction), "effects on expression"))
trans_rates <- trans_rates %>% 
  mutate(effect_direction = paste(str_to_title(effect_direction), "effects on expression"))

# compute average indirect effect rates per effect size direction
avg_trans_rates <- trans_rates %>% 
  group_by(effect_direction) %>% 
  summarize(hit_rate = mean(indirect_rate))

# add rate labels for plots
direct_effects_1mb <- mutate(direct_effects_1mb, rate_label = "Predicted direct effects")
avg_trans_rates <- mutate(avg_trans_rates, rate_label = "Avg. indirect effects rate")
  
# plot predicted effects rate as function of distance to TSS plus average indirect rate
p2 <- ggplot(direct_effects_1mb,
             aes(x = dist_to_tss, y = predicted_direct_rate, color = rate_label)) +
  facet_wrap(~effect_direction, scales = "free_y") +
  geom_line() +
  geom_hline(data = avg_trans_rates, aes(yintercept = hit_rate, color = rate_label)) +
  labs(x = "Distance to TSS (bp)", y = "Effects rate", color = "Effects rate:") +
  scale_color_manual(values = c(`Avg. indirect effects rate` = "red",
                                `Predicted direct effects` = "black")) +
  scale_x_continuous(labels = scales::label_scientific()) +
  scale_y_continuous(limits = c(0, 1)) +
  theme_classic() +
  theme(strip.background = element_blank(), text = element_text(size = 13),
        strip.text = element_text(hjust = 0, size = 16, margin = margin(b = 8)))

p2
```

```{r, fig.height=3, fig.width=8}
# compute average indirect effect rates only for datasets other than DC-TAP-seq
trans_rates_plot <- trans_rates %>% 
  mutate(rate_label = case_when(
    dataset_plots == "DC-TAP-seq (K562)" ~ "K562 indirect effects rate",
    dataset_plots == "DC-TAP-seq (WTC11)" ~ "WTC11 indirect effects rate",
    TRUE ~ "Avg. indirect effects rate other datasets"
  )) %>% 
  group_by(rate_label, effect_direction) %>% 
  summarize(hit_rate = mean(indirect_rate))

# colors for the different rates
rate_colors <- c(
  `Predicted direct effects` = "black",
  `K562 indirect effects rate` = "red",
  `WTC11 indirect effects rate` = "blue",
  `Avg. indirect effects rate other datasets` = "#637a91"
  
)

# plot rates and clip y-axis at 0.1 for better visualization
p3 <- ggplot(direct_effects_1mb,
             aes(x = dist_to_tss, y = predicted_direct_rate, color = rate_label)) +
  facet_wrap(~effect_direction, scales = "free_y") +
  geom_line() +
  geom_hline(data = trans_rates_plot, aes(yintercept = hit_rate, color = rate_label)) +
  labs(x = "Distance to TSS (bp)", y = "Effects rate\n(Displayed range: [0, 0.1])",
       color = "Effects rate:") +
  scale_color_manual(values = rate_colors) +
  scale_x_continuous(labels = scales::label_scientific()) +
  coord_cartesian(ylim = c(0, 0.1)) +
  theme_classic() +
  theme(strip.background = element_blank(), text = element_text(size = 13),
        strip.text = element_text(hjust = 0, size = 16, margin = margin(b = 8)))

p3
```

***

## Arrange supplementary figure
```{r, fig.height=8, fig.width=8}
# arrange panels into figure
plot_grid(
  p1,
  p2 + theme(legend.position = "bottom"),
  labels = c("a", "b"),
  ncol = 1,
  scale = 0.9,
  rel_heights = c(1, 1.1)
  )

# save figure to pdf
ggsave(file = "plots/FigS15_direct_indirect_rates_avg.pdf", height = 8, width = 8, create.dir = TRUE)
```


```{r, fig.height=8, fig.width=8}
# arrange panels into figure
plot_grid(
  p1,
  p3 + theme(legend.position = "bottom") + guides(color = guide_legend(nrow = 2)),
  labels = c("a", "b"),
  ncol = 1,
  scale = 0.9,
  rel_heights = c(1, 1.2)
  )

# save figure to pdf
ggsave(file = "plots/FigS15_direct_indirect_rates_k562_wtc11.pdf", height = 8, width = 8, create.dir = TRUE)
```
***

## Session information
This document was produced using following packages:
```{r sessionInfo}
sessionInfo()
```
